---
title: "Hands-on_Ex03"
author: "TAN Chin Khoon"
date: "12 September 2025"
toc: true
toc-location: right
toc-depth: 3
number-sections: false
---
# 6 Spatio-Temporal Point Patterns Analysis

## 6.1 Overview
A spatio-temporal point process is a random collection of events identified by both time and location, such as disease cases, species sightings, or natural disasters. With the rise of geographically and temporally indexed data, analyzing these patterns has become increasingly important across many fields. In the past decade, several R methods and packages have been developed to support such analyses. This hands-on exercise demonstrates how these tools can be combined in a guided way, using forest fire events in Kepulauan Bangka Belitung, Indonesia (from 1 January to 31 December 2023) as a real-world case study to illustrate the procedures and interpretations.

## 6.2 Learning Outcome

### 6.2.1 The research question
The specific questions we would like to answer are:

- Are the locations of forest fire in Kepulauan Bangka Belitung spatial and spatio-temporally independent?
- If the answer is NO, where and when the observed forest fire locations tend to cluster?

## 6.3 The data
Know what files you’ll use and where they come from.

- forestfires.csv: point events (each row = a fire). Has longitude, latitude, and date/time fields from MODIS.
- Kepulauan_Bangka_Belitung shapefile: the study region polygon (administrative boundary). We’ll read only the Kepulauan Bangka Belitung subset for 2023 analyses.

## 6.4 Installing and Loading the R packages
Load all libraries used in this hands-on exercise.

```{r}
pacman::p_load(sf,           # read, write, and transform spatial vector data
               terra, 
               spatstat, 
               sparr, 
               tmap, 
               tidyverse
               )

```


```{r}



```




```{r}



```




```{r}



```



```{r}



```




```{r}



```




```{r}



```



```{r}



```


```{r}



```



```{r}



```



```{r}



```



```{r}



```



```{r}



```


```{r}



```




```{r}



```



```{r}



```



```{r}



```



```{r}



```



```{r}



```



```{r}



```




```{r}



```



```{r}



```




```{r}



```




```{r}



```




```{r}



```




```{r}



```




```{r}



```




```{r}



```




```{r}



```




```{r}



```



```{r}



```



```{r}



```



```{r}



```



```{r}



```



```{r}



```




```{r}



```





```{r}



```



```{r}



```



```{r}



```




```{r}



```




```{r}



```




```{r}



```




```{r}



```























































































