---
title: "Hands-on Ex08"
author: "TAN Chin Khoon"
date: "24 October 2025"
date-modified: "24 October 2025"
---

# 14 Geographically Weighted Predictive Models

This practical reproduces the complete workflow to build **geographically weighted predictive models** (GWR and Geographically Weighted Random Forest), alongside **non-spatial baselines** (Multiple Linear Regression and Random Forest). Follow each subsection in order.

## 14.1 Overview

Predictive modelling estimates an unknown outcome (e.g., resale price) from known predictors (e.g., floor area, amenities). When observations are georeferenced, relationships may vary across space due to infrastructure, socio-economic and environmental context. Geographically weighted models allow coefficients or model structure to change by location, capturing local effects that global models miss.

By the end, we will:

-   prepare train/test datasets with proper sampling,\
-   check collinearity,\
-   fit and save MLR, GWR, RF, and GW-RF models,\
-   generate out-of-sample predictions,\
-   compute RMSE and visualise prediction quality.

## 14.2 The Data

we will work with:

-   Aspatial table: HDB resale transactions (CSV → converted to sf during preprocessing).\
-   Geospatial layers: URA 2014 Master Plan Planning Subzones (polygon sf).\
-   Locational factors with coordinates: eldercare, hawker centres, parks, supermarkets,\
-   MRT/LRT stations, bus stops, kindergartens, childcare (shapefile/GeoJSON).\
-   Locational factors without coordinates: CBD centroid (derived), shopping malls, primary school rankings (CSV/other).

## 14.3 Installing and Loading R Packages

```{r}

# Create/load all required packages in one shot
pacman::p_load(
  sf,            # spatial vector data (simple features)
  spdep,         # spatial dependence utilities (used by some workflows)
  GWmodel,       # Geographically Weighted Regression
  SpatialML,     # Geographically Weighted Random Forest (grf)
  tmap,          # cartography (not central here but part of the stack)
  rsample,       # train/test splitting (tidymodels)
  Metrics,       # RMSE and other metrics
  tidyverse      # dplyr, ggplot2, readr, purrr, etc.
)

```

## 14.4 Preparing Data

### 14.4.1 Reading data file to rds

```{r}

# Read the prepared modelling dataset (sf object)
mdata <- readr::read_rds("/Users/cktan/Desktop/SMU/01_Geospatial Analytics (ISSS626)/Hands-on_Ex/Hands-on_Ex08/data/rawdata/mdata.rds")

```

### 14.4.2 Data Sampling

The entire data are split into training and test data sets with 65% and 35% respectively by using initial_split() of rsample package. rsample is one of the package of tigymodels.

```{r}

# Set seed to make the split reproducible
set.seed(1234)

# Split into 65% training and 35% testing using rsample
resale_split <- rsample::initial_split(mdata, prop = 6.5/10)

# Extract the two partitions
train_data <- rsample::training(resale_split)
test_data  <- rsample::testing(resale_split)

```

```{r}

# Persist the splits for reuse
readr::write_rds(train_data, "/Users/cktan/Desktop/SMU/01_Geospatial Analytics (ISSS626)/Hands-on_Ex/Hands-on_Ex08/data/model/train_data.rds")
readr::write_rds(test_data,  "/Users/cktan/Desktop/SMU/01_Geospatial Analytics (ISSS626)/Hands-on_Ex/Hands-on_Ex08/data/model/test_data.rds")

```

## 14.5 Computing Correlation Matrix

Before loading the predictors into a predictive model, it is always a good practice to use correlation matrix to examine if there is sign of multicolinearity.

```{r}

# Remove geometry to compute numeric correlations only
mdata_nogeo <- mdata %>% sf::st_drop_geometry()

# Draw an upper-triangle correlation matrix with numbers
corrplot::corrplot(
  cor(mdata_nogeo[, 2:17]), # adjust columns to your numeric predictors
  diag   = FALSE,           # do not draw the diagonal
  order  = "AOE",           # sort for visual clarity
  tl.pos = "td",            # variable labels on top diagonal
  tl.cex = 0.5,             # smaller labels
  method = "number",        # show correlation values
  type   = "upper"          # only the upper triangle
)

```

::: callout-note
If all absolute correlations are below \~0.8, severe multicollinearity is unlikely.
:::

## 14.6 Retrieving the Stored Data

```{r}
# Reload saved splits when resuming work
train_data <- readr::read_rds("/Users/cktan/Desktop/SMU/01_Geospatial Analytics (ISSS626)/Hands-on_Ex/Hands-on_Ex08/data/model/train_data.rds")
test_data  <- readr::read_rds("/Users/cktan/Desktop/SMU/01_Geospatial Analytics (ISSS626)/Hands-on_Ex/Hands-on_Ex08/data/model/test_data.rds")

```

## 14.7 Building a Non-Spatial Multiple Linear Regression

```{r}

# Fit a global (non-spatial) linear regression as a baseline
price_mlr <- lm(
  resale_price ~ floor_area_sqm +
    storey_order + remaining_lease_mths +
    PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
    PROX_MRT + PROX_PARK + PROX_MALL +
    PROX_SUPERMARKET +
    WITHIN_350M_KINDERGARTEN + WITHIN_350M_CHILDCARE +
    WITHIN_350M_BUS + WITHIN_1KM_PRISCH,
  data = train_data
)

# Inspect coefficients and diagnostics
summary(price_mlr)

```

```{r}

# Save the fitted model
readr::write_rds(price_mlr, "/Users/cktan/Desktop/SMU/01_Geospatial Analytics (ISSS626)/Hands-on_Ex/Hands-on_Ex08/data/model/price_mlr.rds")

```

## 14.8 GWR Predictive Method

In this section, you will learn how to calibrate a model to predict HDB resale price by using geographically weighted regression method of GWmodel package.

### 14.8.1 Computing adaptive bandwidth (CV)

```{r}

# Determine optimal adaptive bandwidth (in neighbors) using CV
bw_adaptive <- GWmodel::bw.gwr(
  formula = resale_price ~ floor_area_sqm +
    storey_order + remaining_lease_mths +
    PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
    PROX_MRT + PROX_PARK + PROX_MALL +
    PROX_SUPERMARKET +
    WITHIN_350M_KINDERGARTEN + WITHIN_350M_CHILDCARE +
    WITHIN_350M_BUS + WITHIN_1KM_PRISCH,
  data     = train_data,   # training sf
  approach = "CV",         # cross-validation
  kernel   = "gaussian",   # Gaussian kernel
  adaptive = TRUE,         # adaptive neighbor count
  longlat  = FALSE         # data are in projected meters
)

# Persist the chosen bandwidth (e.g., result may be 40 neighbors)
readr::write_rds(bw_adaptive, "/Users/cktan/Desktop/SMU/01_Geospatial Analytics (ISSS626)/Hands-on_Ex/Hands-on_Ex08/data/model/bw_adaptive.rds")

```

### 14.8.2 Constructing the adaptive bandwidth GWR model

First, let us call the save bandwidth by using the code chunk below.

```{r}

# Reload bandwidth when needed
bw_adaptive <- readr::read_rds("/Users/cktan/Desktop/SMU/01_Geospatial Analytics (ISSS626)/Hands-on_Ex/Hands-on_Ex08/data/model/bw_adaptive.rds")

```

Now, we can go ahead to calibrate the gwr-based hedonic pricing model by using adaptive bandwidth and Gaussian kernel as shown in the code chunk below.

```{r}

# Calibrate GWR using the selected adaptive bandwidth
gwr_adaptive <- GWmodel::gwr.basic(
  formula = resale_price ~ floor_area_sqm +
    storey_order + remaining_lease_mths +
    PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
    PROX_MRT + PROX_PARK + PROX_MALL +
    PROX_SUPERMARKET +
    WITHIN_350M_KINDERGARTEN + WITHIN_350M_CHILDCARE +
    WITHIN_350M_BUS + WITHIN_1KM_PRISCH,
  data     = train_data,    # training sf
  bw       = bw_adaptive,   # adaptive neighbors (numeric)
  kernel   = "gaussian",    # kernel shape
  adaptive = TRUE,          # use adaptive bandwidth
  longlat  = FALSE          # projected coordinates
)

# Save the fitted GWR object
readr::write_rds(gwr_adaptive, "/Users/cktan/Desktop/SMU/01_Geospatial Analytics (ISSS626)/Hands-on_Ex/Hands-on_Ex08/data/model/gwr_adaptive.rds")

```

:::{.callout.tip}

Now, we can go ahead to calibrate the gwr-based hedonic pricing model by using adaptive bandwidth and Gaussian kernel as shown in the code chunk below.

:::




```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```

```{r}



```
