---
title: "Hands-on_Ex02(a)"
author: "TAN Chin Khoon"
date: "`r Sys.Date()`"
---

# 1st Order Spatial Point Patterns Analysis Methods

## 4.3 Installing and Loading the R packages

```{r,}
# Install 'pacman' once if you do not have it (uncomment next line and run only once)
# install.packages("pacman")

# Use pacman::p_load() to install (if needed) and then load the listed packages automatically
pacman::p_load(
  sf,            # modern vector geospatial data handling (Simple Features)
  terra,         # modern raster + vector geospatial data handling
  spatstat,      # meta‑package: tools for spatial point pattern analysis (SPPA)
  tmap,          # cartographic and interactive mapping
  rvest,         # web‑scraping helper used here to parse HTML inside KML attributes
  tidyverse      # data wrangling helpers (dplyr, purrr, stringr, readr, etc.)
)

```

## 4.4 Importing and Wrangling Geospatial Data Sets

Import the Master Plan 2019 Subzone (No Sea) polygons as sf and set projection to EPSG:3414 (SVY21 / Singapore TM)

```{r}
# Read the subzone boundary KML into an sf object named 'mpsz_sf'
mpsz_sf <- sf::st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaKML.kml") %>%
  sf::st_zm(drop = TRUE, what = "ZM") %>%  # remove Z (elevation) and M (measure) dimensions to keep 2D only
  sf::st_transform(crs = 3414)   

```
> **Why remove Z/M?** Our analysis is purely 2D on a planar map. Extra dimensions are unnecessary and can break some operations. \
> **Why EPSG:3414?** This is the standard projected system for Singapore. Distances/areas become meaningful (meters) instead of degrees.

Extract REGION_N, PLN_AREA_N, SUBZONE_N, SUBZONE_C from the Description field (HTML inside KML) (as in the guidance helper function)

```{r}
# Define a small helper function to pull one named item from the HTML table stored in the 'Description' field
extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_) # if Description is empty, return NA immediately

  page <- rvest::read_html(html_text)                              # parse the HTML string
  rows <- rvest::html_elements(page, "tr")                         # get all table rows (<tr>)

  # Find the row whose header cell (<th>) equals the field_name, then take the value in its <td>
  value <- rows %>%
    purrr::keep(~ rvest::html_text2(rvest::html_element(.x, "th")) == field_name) %>%
    rvest::html_element("td") %>%
    rvest::html_text2()

  if (length(value) == 0) NA_character_ else value                 # if not found, return NA
}

```

```{r}
# Apply the function to build clean attributes from 'Description'
mpsz_sf <- mpsz_sf %>%
  dplyr::mutate(
    REGION_N   = purrr::map_chr(Description, extract_kml_field, "REGION_N"),   # region name
    PLN_AREA_N = purrr::map_chr(Description, extract_kml_field, "PLN_AREA_N"), # planning area name
    SUBZONE_N  = purrr::map_chr(Description, extract_kml_field, "SUBZONE_N"),  # subzone name
    SUBZONE_C  = purrr::map_chr(Description, extract_kml_field, "SUBZONE_C")   # subzone code
  ) %>%
  dplyr::select(-Name, -Description) %>%     # drop original noisy fields
  dplyr::relocate(geometry, .after = dplyr::last_col())  # move geometry column to the end for readability

```

> **Checkpoint:** You now have clean polygon attributes and the proper CRS.

(Optional) Filter out offshore/irrelevant areas exactly like in the slides

```{r}
# Create a cleaned version that excludes southern group, western islands, and north‑eastern islands (as per slides)
mpsz_cl <- mpsz_sf %>%
  dplyr::filter(
    SUBZONE_N != "SOUTHERN GROUP",
    PLN_AREA_N != "WESTERN ISLANDS",
    PLN_AREA_N != "NORTH-EASTERN ISLANDS"
  )

# Save a small RDS for repeatability (optional)
# readr::write_rds(mpsz_cl, "chap04/data/mpsz_cl.rds")

```

Import Child Care Services points as sf, drop Z/M, and project to EPSG:3414

```{r}
# Read childcare locations KML into an sf object named 'childcare_sf'
childcare_sf <- sf::st_read("data/geospatial/ChildCareServices.kml") %>%
  sf::st_zm(drop = TRUE, what = "ZM") %>%     # keep 2D only
  sf::st_transform(crs = 3414)         

```
### 4.4.1 Mapping the Geospatial Data Sets

```{r}
# Quick mapping of both layers to verify alignment (childcare points inside planning subzones)
tmap::tmap_mode("plot")  # ensure static plotting mode

# Plot subzones with childcare points overlaid
tmap::tm_shape(mpsz_cl) +
  tmap::tm_polygons(col = "grey85", border.col = "black") +
  tmap::tm_shape(childcare_sf) +
  tmap::tm_dots(size = 0.2, col = "red")

```
> **What to check:** All childcare centres (black dots) fall neatly inside Singapore’s subzones (grey polygons). If not, check CRS and transformations.

Interactive mapping with tmap + leaflet

```{r}
# Switch tmap into interactive mode (internally uses the leaflet library)
tmap::tmap_mode("view")                # enable pan/zoom and feature popups

# Draw an interactive point map on top of a web basemap
# - Click a point to see its attributes
# - Use the layer control to change basemaps (e.g., ESRI.WorldGrayCanvas, OpenStreetMap)
tmap::tm_shape(childcare_sf) +         # select the childcare sf layer
  tmap::tm_dots()                      # render as clickable points

# IMPORTANT: After exploring, switch back to static plotting for the rest of the workflow
# This avoids keeping too many active web-map connections during knitting/deployment
tmap::tmap_mode("plot")               # restore static plotting mode

```


```{r}
# Switch to interactive mode for web-map style exploration
# tmap::tmap_mode("view")
# 
# # Create an interactive map where you can zoom and click points
# tmap::tm_shape(mpsz_cl) +
#   tmap::tm_polygons(col = "lightgrey", border.col = "white") +
#   tmap::tm_shape(childcare_sf) +
#   tmap::tm_dots(col = "blue", size = 0.5)
# 
# # Always switch back to static plotting when done
# tmap::tmap_mode("plot")

```

> **Notes:** In interactive mode you can zoom, pan, and click childcare points to view their attributes. Remember to switch back to plot mode before continuing with static maps.

## 4.5 Geospatial Data Wrangling

### 4.5.1 Convert sf Point Data to ppp (planar point pattern)

```{r}
# Convert the sf point layer to spatstat's planar point pattern (ppp) object
childcare_ppp <- spatstat.geom::as.ppp(childcare_sf)  # now suitable for SPPA functions

```

```{r}
# Verify the class
class(childcare_ppp)  # expect: "ppp"

```

```{r}
# Peek at summary to understand the point pattern and its window
summary(childcare_ppp)  # quick glance (number of points, intensity, window size)

```

### 4.5.2 Create an Owin (observation window) from the Singapore Boundary

```{r}
# Convert the cleaned subzones into a single observation window (owin) for clipping/analysis
sg_owin <- spatstat.geom::as.owin(mpsz_cl)

```

```{r}
# Verify the class
class(sg_owin)  # expect: "owin"

```

```{r}
# Visualize the boundary to confirm it looks right
plot(sg_owin, main = "sg_owin — Singapore Observation Window")

```

## 4.3 Combine the ppp with the owin (keep only points inside Singapore)

```{r}
# Clip the point pattern to the Singapore window to exclude any points outside the boundary
childcareSG_ppp <- childcare_ppp[sg_owin]

```

```{r}
# Confirm the combined object (ppp with polygon window information)
childcareSG_ppp  # should report a polygonal window and the point count

```

## 4.6 Clark-Evan Test for Nearest Neighbour Analysis

**Hypotheses:**

- $H_0$: Childcare centres are randomly distributed (Complete Spatial Randomness, CSR).

- $H_1$: Childcare centres are not randomly distributed (clustered or regular).

The 95% confident interval will be used.

The Clark–Evans test returns an index *R*:

- $R < 1 → clustered$.
- $R = 1 → random$
- $R > 1 → regular$

### 4.6.1 Clark–Evans without CSR Simulation (Z-test)

```{r}
# Run Clark–Evans using observed pattern only (Z-test)
ce_noCSR <- spatstat.explore::clarkevans.test(
X = childcareSG_ppp, # ppp object of childcare centres
correction = "none", # no edge correction (per slides)
alternative = "clustered" # one-sided: test for clustering (R < 1)
)

# Show results
ce_noCSR

```

### 4.6.2 Clark–Evans with CSR via Monte‑Carlo (nsim = 99)

```{r}
# Monte‑Carlo Clark–Evans: compare observed R to R from CSR simulations (nsim = 99)
ce_MC <- spatstat.explore::clarkevans.test( # compute simulated p‑value under CSR
X = childcareSG_ppp, # same ppp as above (metres)
correction = "none", # keep consistent with slides
alternative = "clustered", # one‑sided for clustering (R < 1)
method = "MonteCarlo", # enable Monte‑Carlo simulation mode
nsim = 99 # number of CSR replicates as per slides
) # end clarkevans.test call

ce_MC

```




```{r}


```




```{r}


```



```{r}


```




```{r}


```




```{r}


```


```{r}


```




```{r}


```




```{r}


```



```{r}


```


```{r}


```




```{r}


```




```{r}


```




```{r}


```



```{r}


```




```{r}


```



```{r}


```


































































































































































