---
title: "Hands-on Ex09"
author: "TAN Chin Khoon"
date: "07 November 2025"
date-modified: "last-modified"
---

# 17  Modelling Geographical Accessibility

## 17.1 Introduction

In this exercise you will learn to model geographical accessibility in R using packages for simple features, mapping and accessibility computation. You will prepare the data, compute accessibility with three methods and create maps and boxplots to interpret the results.

### 17.1.1 Learning Outcome

- Import GIS polygon and point data into R as simple feature objects using sf functions

- Import a distance matrix and prepare it for modelling

- Compute accessibility using Hansen, KD2SFCA and Spatial Accessibility Measure methods

- Visualise results with `tmap` and `ggplot2`

## 17.2 The data

Four data sets are used in this exercise, they are:

- `MP14_SUBZONE_NO_SEA_PL`: URA Master Plan 2014 subzone boundary GIS data. This data set is downloaded from data.gov.sg.

- `hexagons`: A 250m radius hexagons GIS data. This data set was created by using st_make_grid() of sf package. It is in ESRI shapefile format.

- `ELDERCARE`: GIS data showing location of eldercare service. This data is downloaded from data.gov.sg. There are two versions. One in ESRI shapefile format. The other one in Google kml file format. For the purpose of this hands-on exercise, ESRI shapefile format is provided.

    - `OD_Matrix`: a distance matrix in csv format. There are six fields in the data file. They are:
    - `origin_id`: the unique id values of the origin (i.e. `fid` of hexagon data set.),
    - `destination_id`: the unique id values of the destination (i.e. `fid` of `ELDERCARE` data set.),
    - `entry_cost`: the perpendicular distance between the origins and the nearest road),
    - `network_cost`: the actual network distance from the origin and destination,
    - `exit_cost`: the perpendicular distance between the destination and the nearest road), and
    - `total_cost`: the summation of `entry_cost`, `network_cost` and `exit_cost`.

:::{.callout-note}

Except `MP14_SUBZONE_NO_SEA_PL` data set, the other three data set are specially prepared by Prof. Kam for teaching and research purpose. Students taking IS415 Geospatial Analytics and Applications are allowed to use them for hands-on exercise purpose. Please obtain formal approval from Prof. Kam if you want to use them for other courses or usage.

:::









```{r}



```



```{r}



```



```{r}



```



```{r}



```




```{r}



```



```{r}



```




```{r}



```




```{r}



```



```{r}



```




```{r}



```




```{r}



```




```{r}



```




```{r}



```




```{r}



```




```{r}



```




```{r}



```




```{r}



```




```{r}



```





```{r}



```




```{r}



```




```{r}



```





```{r}



```




```{r}



```



```{r}



```





```{r}



```



```{r}



```




```{r}



```




```{r}



```




```{r}



```




```{r}



```




```{r}



```




```{r}



```




```{r}



```




```{r}



```




```{r}



```





```{r}



```




```{r}



```




```{r}



```




```{r}



```



```{r}



```




```{r}



```




```{r}



```




```{r}



```




```{r}



```




```{r}



```




```{r}



```




```{r}



```




```{r}



```



```{r}



```





```{r}



```





```{r}



```





```{r}



```





```{r}



```




```{r}



```




```{r}



```




```{r}



```





```{r}



```



```{r}



```





```{r}



```





```{r}



```




```{r}



```





```{r}



```




```{r}



```




```{r}



```





```{r}



```





```{r}



```



```{r}



```




```{r}



```




```{r}



```




```{r}



```




```{r}



```




```{r}



```




```{r}



```





```{r}



```




```{r}



```




```{r}



```




```{r}



```




```{r}



```




```{r}



```



```{r}



```



```{r}



```




```{r}



```




```{r}



```




```{r}



```





```{r}



```




```{r}



```




```{r}



```




```{r}



```




```{r}



```




```{r}



```




```{r}



```



















































































